---
title: "Descriptive Statistics And Graphs"
author: "Jie Wang"
date: "2025-05-02"
output: html_document
---

This file contains sample R code for the lesson "Descriptive Statistics and Graphs".

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example 1

Let's use a toy example. For more details, please read the lesson notes. 

```{r race0}
Time = c(44, 37, 40, 46, 44, 30, 40, 34, 34, 37, 
        36, 42, 39, 36, 37, 37, 42, 34, 37, 31)
Time
```
Let's find the mean and standard deviation of Time. 

```{r race1}
time.bar = mean(Time)
time.sd <- sd(Time)

```

The (sample) mean time is `r round(time.bar,3)` with a (sample) standard deviation `r round(time.sd, 3)`. 

If you would like to obtain a set of summary statistics, you may try the following:

```{r race2}

# The five number summary and the mean
summary(Time)

# Range
range(Time)

# IQR
IQR(Time)

# Variance
var(Time)

```
Let's get the graphs. 

```{r race3}

# Histogram
hist(Time, breaks=seq(30,46,2), right = FALSE) 
# right= is logical; if TRUE, the histogram cells are right-closed (left open) intervals.

# Boxplot
boxplot(Time, xlab="Time", ylab="Minutes")

```
## Example 2

Let's see another type of variable. 

```{r food}
Food = c('Pinapple', 'Pinapple', 'Pasta', 'Pizza', 'Pasta', 'Pasta', 'Pinapple', 'Pizza', 'Pasta', 'Pasta', 'Pinapple', 'Pasta', 'Pizza', 'Pinapple', 'Pasta', 'Pizza', 'Pinapple', 'Pinapple', 'Pinapple', 'Pinapple')
print(Food)
```


Obtain the frequency (count) or relative frequency of each type of food. 

```{r food1}
table(Food)

table(Food)/length(Food)
```
Making graphs. 

```{r food2}

Freq <- c(7,9,4)
FoodType <- c("Pasta","Pinapple","Pizza")

# Bar chart. You can add a lot of attributes. 
barplot(Freq,names.arg=FoodType,xlab="Food",ylab="Frequency",
        col="blue", main="Bar Chart for Food Type",border="red")

# Pie chart. You can make it very simple. 
pie(Freq, FoodType)

```


## Example 3 

```{r cars}
data(mtcars)
mtcars

# Or you can print just a few rows. 
head(mtcars)

```

Now, let's play the game of finding statistics and making graphs using this set of data.

Can you find the descriptive statistics of the variable mpg?

Can you find the descriptive statistics of cyl, treating cyl as a categorical variable?

Can you make a histogram for mpg? a bar chart for cyl?

Let me show you what I do to answer these questions. 


```{r cars1}

# The (sample) mean of mpg
mpg.bar = mean(mtcars$mpg) # Use the $ sign to reference a variable in a data frame. 
mpg.bar

# If you don't want to use the $ sign, you can try with() as shown below. 
# The sample standard deviation of mpg
mpg.sd = with(mtcars, sd(mpg))
mpg.sd

# The sample variance can be found by 
var(mtcars$mpg)
# or 
mpg.sd^2

# The quartiles and the extremes can be found individually or together as shown below. 
quantile(mtcars$mpg)
```

Thus the mean and standard deviation of mpg are `r round(mpg.bar,3)` and `r round(mpg.sd,3)`. 

```{r cars2}

# The frequencies of cyl values
table(mtcars$cyl)

# Relative frequencies
table(mtcars$cyl)/length(mtcars$cyl)

```
```{r cars3}

# The histogram
hist(mtcars$mpg)

# The bar chart
Freq=c(11,7,14)
cyl_cate=c("4","6","8")

barplot(Freq,names.arg=cyl_cate)
```